Type Refinements for the Working Class

### 2015-12-31 paper.tex
finish up definition of abts (close #14)

Fix a set of \emph{sorts} $\Sorts$; we'll say $\IsSort{\tau}$ when
$\Member{\tau}{\Sorts}$. Let $\SymSets$ be the category of finite cardinals and
injective maps between them; using the comma construction, we can define the
category $\SCtx$ of \emph{contexts} of symbols, whose objects are finite sets
of symbols $U$ and sort-assignments
\begin{tikzcd}[cramped]
  \IMode{U}\arrow[r, "\IMode{\mathfrak{s}}"] &\IMode{\Sorts}
\end{tikzcd}, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.

We will say that an $\Sorts$-valence is a form
$\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}$, where
$\IMode{\vec{\sigma}\cup\vec{\tau}}\subseteq\IMode{\Sorts}$ and
$\IsSort{\tau}$; valences represent the \emph{binding structure} of an
abstraction of sort $\tau$, where $\vec{\sigma}$ are the sorts of the bound
symbols and $\vec{\tau}$ are the sorts of the bound variables. When $v$ is a
valence, we will say $\IsValence{v}$. An $\Sorts$-arity is of the form
$\MkArity{\vec{v}}{\tau}$, where $\vec{v}$ are $\Sorts$-valences and
$\IsSort{\tau}$; arities classify \emph{operators} (also called ``function
symbols'' in older treatments of algebra). When $a$ is an arity, we will say
$\IsArity{a}$.

An \emph{operator signature} is an arity-indexed family of copresheaves on
$\SCtx$, namely $\HypJ{\Of{\Operators_a}{\Sets^\SCtx}}{\IsArity{a}}$; this is
an intensional generalization of the state of affairs for standard universal
algebra, where operators are arity-indexed families of \emph{sets}. The point
of fibering over symbol contexts is to introduce an open-ended indexing
mechanism for operators; the functorial liftings of $\SCtx$-morphisms
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
equip operators $\Member{\vartheta}{\Operators_a(\Upsilon)}$ with a suitable
renaming action $\Member{\Rename{\rho}{\vartheta}}{\Operators_a(\Upsilon')}$.%
%
\footnote{
  In~\cite{sterling-morrison:2015}, the authors defined operator signatures as
  \emph{sheaves} on $\OpCat{(\SCtx)}$ under the atomic topology (alternatively,
  pullback-preserving presheaves), where all non-empty sieves cover; the sheaf
  condition induces the unique existence of \emph{strengthenings} for
  operators. However, because we do not use strengthening in the present
  development, we have omitted this part of the apparatus for the sake of
  simplicity.
}

% \todo[inline]{give a brief description of the framework, and present its rules.}

>>>>

Fix a set of \emph{sorts} $\Sorts$; we'll say $\IsSort{\tau}$ when
$\Member{\tau}{\Sorts}$. Let $\SymSets$ be the category of finite cardinals and
injective maps between them; using the comma construction, we can define the
category $\SCtx$ of \emph{contexts} of symbols, whose objects are finite sets
of symbols $U$ and sort-assignments
\begin{tikzcd}[cramped]
  \IMode{U}\arrow[r, "\IMode{\mathfrak{s}}"] &\IMode{\Sorts}
\end{tikzcd}, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.  We also write $\IsSymCtx{\Upsilon}$ when $\Member{\Upsilon}{\SCtx}$.

We will say that an $\Sorts$-valence is a form
$\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}$, where
$\IMode{\vec{\sigma}\cup\vec{\tau}}\subseteq\IMode{\Sorts}$ and
$\IsSort{\tau}$; valences represent the \emph{binding structure} of an
abstraction of sort $\tau$, where $\vec{\sigma}$ are the sorts of the bound
symbols and $\vec{\tau}$ are the sorts of the bound variables. When $v$ is a
valence, we will say $\IsValence{v}$; let $\Valences$ be the set of $\Sorts$-valences.

An $\Sorts$-arity is of the form $\MkArity{\vec{v}}{\tau}$, where $\vec{v}$ are
$\Sorts$-valences and $\IsSort{\tau}$; arities classify \emph{operators} (also
called ``function symbols'' in older treatments of algebra). When $a$ is an
arity, we will say $\IsArity{a}$; let $\Arities$ be the set of $\Sorts$-arities.

An $\Sorts$-\emph{operator signature} is an $\Sorts$-arity-indexed family of copresheaves on
$\SCtx$, namely $\HypJ{\Of{\Operators_a}{\Sets^\SCtx}}{\IsArity{a}}$; this is
an intensional generalization of the state of affairs for standard universal
algebra, where operators are arity-indexed families of \emph{sets}. The point
of fibering over symbol contexts is to introduce an open-ended indexing
mechanism for operators; the functorial liftings of $\SCtx$-morphisms
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
equip operators $\Member{\vartheta}{\Operators_a(\Upsilon)}$ with a suitable
renaming action $\Member{\Rename{\rho}{\vartheta}}{\Operators_a(\Upsilon')}$.%
%
\footnote{
  In~\cite{sterling-morrison:2015}, the authors defined operator signatures as
  \emph{sheaves} on $\OpCat{(\SCtx)}$ under the atomic topology (alternatively,
  pullback-preserving presheaves), where all non-empty sieves cover; the sheaf
  condition induces the unique existence of \emph{strengthenings} for
  operators. However, because we do not use strengthening in the present
  development, we have omitted this part of the apparatus for the sake of
  simplicity.
}
%
We call the pair $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ an \emph{abt signature}.

Just as we have defined symbol contexts as $\SCtx$, we will use the category
$\FinSets$ of finite cardinals and \emph{all} maps between them to define
variable contexts $\Ctx$ and metavariable contexts $\MCtx$, writing
$\IsVarCtx{\Gamma}$ when $\Member{\Gamma}{\Ctx}$ and $\IsMetaCtx{\Theta}$ when
$\Member{\Theta}{\MCtx}$. The difference between $\SCtx$ and $\Ctx$ is that the
latter allows renamings that may identify distinct names, whereas renamings in
$\SCtx$ are always injective.

For a given abt signature, the syntax of abts is given by the judgments
$\IsAbt{\Theta}{\Upsilon}{\Gamma}{M}{\tau}$ (presupposing $\IsMetaCtx{\Theta}$,
$\IsSymCtx{\Upsilon}$, $\IsVarCtx{\Gamma}$ and $\IsSort{\tau}$) and
$\IsAbs{\Theta}{\Upsilon}{\Gamma}{E}{v}$ (presupposing $\IsMetaCtx{\Theta}$,
$\IsSymCtx{\Upsilon}$, $\IsVarCtx{\Gamma}$ and $\IsValence{v}$), defined in
mutual induction as follows:

\[
  \infer[\RuleIsAbtVar]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{x}{\tau}
  }{
    \Lookup{\Gamma}{x}{\tau}
  }
\]
\[
  \infer[\RuleIsAbtMVar]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Theta}{\MV{m}}{
        \MkValence{
          \sigma_0,\dots,\sigma_m
        }{
          \tau_0,\dots,\tau_n
        }{\tau}
      }
\\
      \HypJ{
        \Lookup{\Upsilon}{u_i}{\sigma_i}
      }{\Leq{i}{m}}
\\
      \HypJ{
        \IsAbt{\Theta}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\]
\[
  \infer[\RuleIsAbtApp]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{v_1,\dots,v_n}{\tau}
      }
\\
      \HypJ{
        \IsAbs{\Theta}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\]
\[
  \infer[\RuleIsAbtAbs]{
    \IsAbs{\Theta}{\Upsilon}{\Gamma}{
      \MkAbs{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsAbt{
      \Theta
    }{
      \Upsilon, \vec{u}:\vec{\sigma}
    }{
      \Gamma,\vec{x}:{\vec{\tau}}
    }{
      M
    }{\tau}
  }
\]

We have several meta-operations on abts, including the calculation of free
symbols $\FS{M}$, the calculation of free variables $\FV{M}$, the transport
$\Rename{\rho}{M}$ across a renaming $\rho$, the substitution $\Subst{N}{x}{M}$
of a term $N$ for the variable $x$ in $M$, and the substitution
$\Subst{E}{\MV{m}}{M}$ of an abstraction $E$ for the metavariable $\MV{m}$ in
$M$. Please see~\cite{sterling-morrison:2015} for the explicit definition of
these operations.
